---
title: "ESM 244 Lab 10 - building versatility"
subtitle: "SQL and Python in R"
author: "Allison Horst"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)

# Packages needed: 
library(tidyverse)
library(here)
library(janitor)

# For working with SQLite databases: 
library(DBI)
library(RSQLite)

# For interfacing between R and Python:
library(reticulate)
```

## 1. SQL in R Markdown with `DBI` and `RSQLite`

Here, we'll explore and query a database containing information on fish, invertebrates, and lobsters from the Santa Barbara Coastal Long Term Ecological Research Program. 

#### Data:

Fish data: 

- Description: Reef fish abundance, SB coast
- Link: https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-sbc&identifier=17&revision=newest
- Citation: Reed D. 2018. SBC LTER: Reef: Kelp Forest Community Dynamics: Fish abundance. Environmental Data Initiative. doi.

##### Invertebrate data:

- Description: Invertebrate counts, SB coast
- Link: https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-sbc&identifier=19&revision=newest
- Citation: Reed D. 2018. SBC LTER: Reef: Kelp Forest Community Dynamics: Invertebrate and algal density. Environmental Data Initiative. doi.

##### Lobster data: 

- Description: Lobster size, abundance and fishing pressure (SB coast)
- Link: https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-sbc&identifier=77&revision=newest
- Citation: Reed D. 2019. SBC LTER: Reef: Abundance, size and fishing effort for California Spiny Lobster (Panulirus interruptus), ongoing since 2012. Environmental Data Initiative. doi.

You should have the `marine.sqlite` in the `data` folder of your project. 

#### a. Connect to the database using `dbConnect()`:
```{r}
# Then connect to the database:
lter_db <- DBI::dbConnect(drv = RSQLite::SQLite(), 
                          dbname = here("data", "marine.sqlite"))
```

From [Data Carpentry](https://datacarpentry.org/R-ecology-lesson/05-r-and-databases.html): "This command uses 2 packages that helps dbplyr and dplyr talk to the SQLite database. DBI is not something that you’ll use directly as a user. It allows R to send commands to databases irrespective of the database management system used. The RSQLite package allows R to interface with SQLite databases."

OK but how do we see what it contains? `View`? What is happening here? 

#### b. See what tables exist in the database:
```{r}
DBI::dbListTables(lter_db)
```

So, this relational database contains 3 tables, named fish_data, inverts_data, and lobsters_data. 

#### c. See what variables exist in the table with `DBI::dbListFields()`:
```{r}
DBI::dbListFields(lter_db, "fish_data")
```

Your turn: what variables exist in the inverts_data and lobsters_data tables?

#### d. Cool! Use `DBI::dbReadTable()` to get a table as a data frame:
```{r}
fish_df <- DBI::dbReadTable(lter_db, "fish_data")

# Check the class:
class(fish_df) # oooo a data frame!
```

Use our usual exploratory tools `View()`, `head()`, etc. to check it out!
Great, a data frame. 

Mmmmm but if things are extra giant then maybe we don't want to read in the *entire* table. Let's do some SQL querying! 

#### e. Query a database with SQL

**SQL: Structured Query Language**

From http://www.sqlcourse.com/intro.html:
"SQL (pronounced "ess-que-el") stands for Structured Query Language. SQL is used to communicate with a database. According to ANSI (American National Standards Institute), it is the standard language for relational database management systems. SQL statements are used to perform tasks such as update data on a database, or retrieve data from a database."

Some important (+ readable, thanks Hadley Wickham) stuff: https://cran.r-project.org/web/packages/RSQLite/vignettes/RSQLite.html

To work in SQL within R Markdown, we have options. First, we can insert a SQL code chunk (instead of a normal R code chunk, which expects R language). 

Use Insert > SQL to create one, and your new code chunk should have the header `{sql connection= }`. Add the name of the database connection you created above (in our case, that will be `sql connection=fish_db` within the code chunk header).

We will use: 

- `SELECT` to get rows 
- `FROM` a table 
- Specifying a subset of variables (columns), or `*` to get all columns

```{sql connection=lter_db}
SELECT * FROM fish_data
```

**Note**: If you have a bunch of SQL code chunks, you might want to set up a connection in the R setup code chunk at the top of the .Rmd. Ours would look like: 
```{r, eval = FALSE}
lter_db <- dbConnect(RSQLite::SQLite(), dbname = "marine.sqlite")
knitr::opts_chunk$set(connection = "lter_db")
```

Now let's just get a few of the columns:
```{sql connection=lter_db}
SELECT year, total_count, site FROM fish_data
```
Let's think about the patterns here:

- The CAPITAL WORDS are like our functions (but actually these aren't case sensitive...I just like the delineation)
- The words between them are like arguments

What if we only want to get observations from the **fish_data** table where the year is 2016? Use `WHERE` to set conditions on what to keep:  
```{sql connection=lter_db}
SELECT * FROM fish_data WHERE year == 2016
```

So what is the parallel function that we've used in R to do the same thing? Yep, dplyr::filter() allows us to conditionally subset by rows! 

We can similarly look for conditional matches on strings (e.g. "garibaldi"):
```{sql connection=lter_db}
SELECT * FROM fish_data WHERE common_name == "garibaldi"
```

What about an **and** statement (we use a comma or ampersand in a `dplyr::filter()` function): 
```{sql connection=lter_db}
SELECT * FROM fish_data WHERE common_name == "garibaldi" and site == "mohk"
```

An **or** statement: 
Keep rows where the year is 2017 **OR** the site is "abur"
```{sql connection=lter_db}
SELECT * FROM fish_data WHERE year == 2017 or site == "abur"
```

If we want to store the output of SQL querying as a data frame, add `output.var = "df_name"` to the code chunk header! 
```{sql connection=lter_db, output.var = "fish_2017_df"}
SELECT * FROM fish_data WHERE year == 2017
```

Then check the class of fish_2017_df:
```{r}
# class(fish_2017_df) # A data frame wooo! 
```

And you actually don't **HAVE** to do this in SQL code chunks...you can use `DBI::dbSendQuery` in an R code chunk to do the same thing!

```{r}
# Now I'm in an R code chunk, querying with SQL!
garibaldi <- dbSendQuery(lter_db, 
                         "SELECT * FROM fish_data WHERE common_name == 'garibaldi'")

# And see what it contains (coerce to a data.frame):
gar_df <- dbFetch(garibaldi)

```

#### f. `dbplyr` for `dplyr`-like querying!

Let's say we want to do something with a database table, but we really enjoy working with `dplyr` type syntax instead. We'll use `dbplyr` to help us out. 

From https://dbplyr.tidyverse.org/: "Note that you don’t actually need to load dbplyr with library(dbplyr); dplyr automatically loads it for you when it sees you working with a database. Database connections are coordinated by the DBI package. Learn more at http://dbi.r-dbi.org/"

```{r}

# Convert to a SQL table:
fish_tbl <- tbl(lter_db, "fish_data")

# Do some wrangling in dplyr format: 
yearly_fish <- fish_tbl %>% 
  group_by(year) %>% 
  summarize(
    total_fish = sum(total_count)
  )

# See query:
yearly_fish %>% show_query() # This will show us what the SQL commands would look like! So cool! 

# Actually run the query by passing it to collect(): 
yearly_fish %>% collect()

# Then you don't even need to WRITE any SQL code! Thanks dbplyr!

```

#### g. Adding / removing tables in a .sqlite database

What if we want to add a table (e.g. from a data frame or CSV) to an existing database? 

Use `dbWriteTable` to add a new table to an existing database. For example, let's add the `mtcars` dataset to our database, then remove it:  
```{r}
# View(mtcars)

# Write the table to our database:
dbWriteTable(lter_db, "mtcars_df", mtcars)

# Check to see that it's there: 
DBI::dbListTables(lter_db) # Yup. 

# And we can remove it just as easily: 
DBI::dbRemoveTable(lter_db, "mtcars_df")

# Now check: 
DBI::dbListTables(lter_db) # Back to the 3 we started with! 

```

## End SQL in R introduction
