---
title: "Python in R with reticulate"
author: "Allison Horst"
output: html_document
---

## Python in R (Markdown) with `reticulate`

Hi 244ers! This part of the lab shows some basic tools for working with Python in R Markdown, using the `reticulate` package. 

### Background: 

The [`reticulate`](https://rstudio.github.io/reticulate/) package provides an "R interface to Python modules, classes, and functions. When calling into Python R data types are automatically converted to their equivalent Python types. When values are returned from Python to R they are converted back to R types. The reticulate package is compatible with all versions of Python >= 2.7."

- **Note 1: If you want to actually do this part of the lab, you will need to do some installation & configuration to get set up on your computer.** I recommend following along [here (through Step 5)](https://docs.rstudio.com/tutorials/user/using-python-with-rstudio-and-reticulate/) to install Python and get up and running with a virtual environment so that you can Py in R too. 

Or just check out the code below for some basics: 

#### a. Create a new .Rmd and attach packages

First, create a new R Markdown document. Then attach the packages we'll need for this part of the lab: 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
library(here)
library(reticulate)
```

#### b. Check python configuration

Yours may be different from mine. 
```{r}
py_config()
```

You could also tell R which Python to use (if needed), but this tends to get not fun really quickly re: installing new libraries, etc. Highly recommend the tutorial up top ^^. 

**Note:** Once you set up your virtual environmental by following the path above, you shouldn't need to do this (but can, if you want to use a different Python version that exists on your computer)
```{r}
# use_python("path/to/python")
# py_config()
```

#### c. Install Python libraries 

Like we install R packages that contain functions or tools we want to use, we might also need to install Python libraries. Check out: https://rstudio.github.io/reticulate/index.html for a lot of examples and information on how you can do that - or use the `reticulate::py_install("library_name")` function for R-like installation syntax!

For example, if I wanted to install SciPy (I'm not actually going to do this here because I have SciPy installed already, but you could uncomment & run this to install if you're set up & configured): 
```{r}
# py_install("scipy")
```

#### d. Just generally working in a Py code chunk

First, just proof that you can work in a Python code chunk in R Markdown.

Click Insert > Python, and you'll see a new code chunk that has a {python} header. That means that code chunk will be expecting Python code. 

We can use it as a basic editor, for example: 

```{python}
a = 4
b = 10
c = a + b
c # This will return c
```

How can I call those in the Console, or later on in an R code chunk, which is expecting R? use `py$object_name` to make sure that R knows you're looking for a Python variable. 

For example, after running the code chunk above,
type `py$c` into the R Console and see that it returns the value of that python object! 

#### e. Import Python libraries

In R, we *install* packages, but also have to *attach* them when we want them available for use (with `library(package_name)`). Similarly, we need to *import* Python packages that we've installed so that we can use them. 

The Python libraries (that I have already installed) I'll use are:

- `pandas`: for exploring & wrangling data frames (also using tidy structure)
- `matplotlib`: is a graphics library
- `dplython`: functions for wrangling that are similar to `dplyr` in R

Here, I given pandas and matplotlib.pyplot (a module in matplotlib) shorter "nicknames" to refer to when I use it later on, but don't have to. 
```{python}
# Attach modules
import pandas as pd
import matplotlib.pyplot as mp

# Get specific pieces from dplython
from dplython import (DplyFrame, X, head, arrange, mutate, select, group_by, summarize)
  
```

#### g. Reading in and exploring data

Comparison of common functions: 
https://pandas.pydata.org/docs/getting_started/comparison/comparison_with_r.html

Starting from `my_data.csv`

- Do some exploring first with R code (View, head, tail, names, dim)
```{r}

df_r <- read_csv(here("data", "my_data.csv"))

View(df_r) # Return entire data frame
head(df_r) # Return first 6 lines
tail(df_r) # Return last 6 lines
names(df_r) # Return variable names
dim(df_r) # Return dimensions

```

Let's do the same with Python code!

First: read it in (`pandas.read_csv`)
```{python}
df_py = DplyFrame(pd.read_csv("../data/my_data.csv"))
```

Then look at it using similar functions: 
```{python}
print(df_py) # View it
```

```{python}
df_py.head(3) # Return the first 3 rows
```

```{python}
df_py.tail(2) # Return the last 2 rows
```

```{python}
list(df_py) # Return the column names
````

```{python}
df_py.shape # Returns dimensions
```

#### h. Wrangling similarities between R and Python 

Now let's do some wrangling and viz, first in R, then showing how to do the same thing in Python. Here, starting from df_r or df_py, we'll: 

- Only keep observations where weight is greater than 12 pounds
- Just keep columns species, weight_lb and height_m
- Add a new column for the weight/height ratio
- Make a scatterplot

```{r}
# Do the wrangling:
new_r <- df_r %>% 
  filter(weight_lb > 12) %>% 
  select(species, weight_lb, height_m) %>% 
  mutate(ratio = weight_lb / height_m)

# Then make a graph with ggplot2:
ggplot(data = new_r, aes(x = weight_lb, y = height_m)) +
  geom_point(size = 4)
```

Let's do the same in Python
Note: `>>` operator is in `dplython` module
```{python}
new_py = df_py >> select(X.weight_lb, X.height_m) >> X.query('weight_lb > 12') >> X.assign(ratio = X['weight_lb'] / X['height_m'])

print(new_py)

# So it's the same thing, and we can plot with matplotlib
```

```{python}
mp.scatter('weight_lb', 'height_m', data = new_py)
```

But also, what if I just want to get this to be an R data frame? No problem! 
```{r}
my_df <- py$df_py
class(my_df)
```

We have seen a number of different ways to use Python in RStudio, and with R code, including: 

1. Run Python scripts (.py) in R Studio
2. Write Python code in R Markdown thanks to `reticulate` 
3. Call Python objects (e.g. pandas DataFrames) in R code chunks with `py$object_name`

