---
title: "ESM 244 Lab 10 - building versatility"
subtitle: "SQL and Python in R"
author: "Allison Horst"
output: html_document
---

### Introduction - Lab 10

Hi everyone! Welcome to Lab 10 - I'll miss seeing you all in person this week, but will try to include everything I say during the lab in this document. In lecture we talked about several other languages (Python, SQL) that are often used in data science and highly desirable for many employers. 

In this document (lab_10_key.Rmd) you can follow along to use SQL to query a relational database. The additional part (reticulate_example.Rmd) demos how to interface between R and Python in R Markdown, but *does* require that you have Python installed and successfully configured, with some Python modules (like R packages) installed. You are **not** required to do that, but if it sounds interesting to you, I recommend getting set up by following along with [this tutorial from RStudio](https://docs.rstudio.com/tutorials/user/using-python-with-rstudio-and-reticulate/).

OK, let's get started with the SQL part of Lab 10.

### SQL (structured query language) to query databases in R

In [Lecture 15](https://docs.google.com/presentation/d/1JBeZ3p7amck8a8jS8xZQwnpdSN_LlXQfP4Q5VnetM4w/edit?usp=sharing) we talked about relational databases, and how we can use SQL to query them. A relational database is a collection of tables containing information, often that are related to each other (e.g. by connected identifiers). Here, we'll learn to:

- Use `DBI` and `RSQLite` to connect to a database
- Explore the tables and variables in the database tables
- Query tables using SQL
- Add a table to a .sqlite database

#### A. Getting started...

For this part of the lab, we'll use 4 packages: 

- `tidyverse`: you know this one
- `here`: better file paths 
- `DBI`: "Defines and interface for communication between R and relational database management systems"
- `RSQLite`: Allows you to connect to a SQLite database file

##### a. Create a new R Markdown document

If you haven't already, open a new R Markdown document and delete everything below the first code chunk. 

##### b. Update the setup code chunk

In the setup code chunk, update the chunk options (if you want) with `warning = FALSE` and `message = FALSE`, then attach the `tidyverse`, `here`, `DBI`, and `RSQLite` packages. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)

# Packages needed: 
library(tidyverse)
library(here)

# For working with SQLite databases: 
library(DBI)
library(RSQLite)
```

Run the code chunk to make sure that they all attach successfully. Cool! Ready to get started with RSQLite. 

#### B. Exploring a .sqlite database

Here, we'll explore and query a database containing information on fish, invertebrates, and lobsters from the Santa Barbara Coastal Long Term Ecological Research Program. 

**Data descriptions:**

Fish data: 

- Description: Reef fish abundance, SB coast
- Link: https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-sbc&identifier=17&revision=newest
- Citation: Reed D. 2018. SBC LTER: Reef: Kelp Forest Community Dynamics: Fish abundance. Environmental Data Initiative. doi.

Invertebrate data:

- Description: Invertebrate counts, SB coast
- Link: https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-sbc&identifier=19&revision=newest
- Citation: Reed D. 2018. SBC LTER: Reef: Kelp Forest Community Dynamics: Invertebrate and algal density. Environmental Data Initiative. doi.

Lobster data: 

- Description: Lobster size, abundance and fishing pressure (SB coast)
- Link: https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-sbc&identifier=77&revision=newest
- Citation: Reed D. 2019. SBC LTER: Reef: Abundance, size and fishing effort for California Spiny Lobster (Panulirus interruptus), ongoing since 2012. Environmental Data Initiative. doi.

You should have the `marine.sqlite` in the `data/` folder of your project. 

##### a. Connect to the database using `dbConnect()`

First, we need to establish a connection with our database (marine.sqlite), and specify which database engine we'll use to work with it (in our case, SQLite). 

We'll store that database connection as 'lter_db', and connect as follows:

```{r}
# Then connect to the database:
lter_db <- DBI::dbConnect(drv = RSQLite::SQLite(), 
                          dbname = here("data", "marine.sqlite"))
```

From [Data Carpentry](https://datacarpentry.org/R-ecology-lesson/05-r-and-databases.html): "This command uses 2 packages that helps dbplyr and dplyr talk to the SQLite database. DBI is not something that you’ll use directly as a user. It allows R to send commands to databases irrespective of the database management system used. The RSQLite package allows R to interface with SQLite databases."

OK but how do we see what the database `lter_db` contains? Unlike with a data frame, we can't use `View()` because this is a database containing multiple tables. We'll use functions from `DBI` to go exploring.

##### b. See what tables exist in the database:

Use `DBI::dbListTables()` to return the names of all tables in a database.

```{r}
DBI::dbListTables(lter_db)
```

So, this relational database contains 3 tables, named: fish_data, inverts_data, and lobsters_data. 

##### c. See what variables exist in a database table

Next, now that we know which *tables* exist in the database, we probably want to know what variables it contains. Use `DBI::dbListFields()` to return the variable (column) names for a table in the database. In this example, we want to return the column names from the *fish_data* table in the lter_db database we're connected to. 

```{r}
DBI::dbListFields(lter_db, "fish_data")
```

When you run that code, you should see that there are four columns: *year*, *site*, *common_name*, and *total_count*

**Activity:** What variables exist in the inverts_data and lobsters_data tables in the lter_db database?

##### d.`DBI::dbReadTable()` to get a table as an R data frame

We may want to read in an entire table from a database, and store it as a data frame. Use `DBI::dbReadTable()` to store a database table as an R data frame object: 

```{r}
fish_df <- DBI::dbReadTable(lter_db, "fish_data")

# Check the class:
class(fish_df) # oooo a data frame!
```

Now that `fish_df` is stored as a data frame, you can use the functions we've been using all along to work with it. Explore `fish_df` using our standards (e.g. `View()`, `head()`, `dim()`, `summary()`).  

In that example, we stored an entire database table as a data frame. But we may want to query from a database table instead of reading in the entire thing. We'll use SQL in a couple different ways to help us do that.  

#### e. Query a database with SQL

**SQL: Structured Query Language**

From http://www.sqlcourse.com/intro.html:
"SQL (pronounced "ess-que-el") stands for Structured Query Language. SQL is used to communicate with a database. According to ANSI (American National Standards Institute), it is the standard language for relational database management systems. SQL statements are used to perform tasks such as update data on a database, or retrieve data from a database."

Some important (+ readable, thanks Hadley Wickham) stuff about `RSQLite`: https://cran.r-project.org/web/packages/RSQLite/vignettes/RSQLite.html

To work in SQL within R Markdown, we have options. First, we can insert a SQL code chunk (instead of a normal R code chunk, which expects R language). 

Use Insert > SQL to create one, and your new code chunk should have the header `{sql connection= }`. Add the name of the database connection you created above (in our case, that will be `sql connection=lter_db` within the code chunk header).

We will use: 

- `SELECT` to get columns 
- `FROM` a table 
- Specifying a subset of variables (columns), or `*` to get all columns

```{sql connection=lter_db}
SELECT * FROM fish_data
```

**Note**: If you have a bunch of SQL code chunks, you might want to set up a connection in the R setup code chunk at the top of the .Rmd. Ours would look like:

    `# lter_db <- dbConnect(RSQLite::SQLite(), dbname = "marine.sqlite")`
    `# knitr::opts_chunk$set(connection = "lter_db")`

Now let's just get a few of the columns:
```{sql connection=lter_db}
SELECT year, total_count, site FROM fish_data
```
Let's think about the patterns here:

- The CAPITAL WORDS are like our functions (but actually these aren't case sensitive...I just like the delineation)
- The words between them are like arguments

What if we only want to get observations from the **fish_data** table where the year is 2016? Use `WHERE` to set conditions on what to keep:  
```{sql connection=lter_db}
SELECT * FROM fish_data WHERE year == 2016
```

So what is the parallel function that we've used in R to do the same thing? Yep, dplyr::filter() allows us to conditionally subset by rows! 

We can similarly look for conditional matches on strings (e.g. "garibaldi"):
```{sql connection=lter_db}
SELECT * FROM fish_data WHERE common_name == "garibaldi"
```

What about an **and** statement (we use a comma or ampersand in a `dplyr::filter()` function): 
```{sql connection=lter_db}
SELECT * FROM fish_data WHERE common_name == "garibaldi" and site == "mohk"
```

An **or** statement: 
Keep rows where the year is 2017 **OR** the site is "abur"
```{sql connection=lter_db}
SELECT * FROM fish_data WHERE year == 2017 or site == "abur"
```

If we want to store the output of SQL querying as a data frame, add `output.var = "df_name"` to the code chunk header! 
```{sql connection=lter_db, output.var = "fish_2017_df"}
SELECT * FROM fish_data WHERE year == 2017
```

Then check the class of fish_2017_df:
```{r}
# class(fish_2017_df) # A data frame wooo! 
```

And you actually don't **HAVE** to do this in SQL code chunks...you can use `DBI::dbSendQuery` in an R code chunk to do the same thing!

```{r}
# Now I'm in an R code chunk, querying with SQL!
garibaldi <- dbSendQuery(lter_db, 
                         "SELECT * FROM fish_data WHERE common_name == 'garibaldi'")

# And see what it contains (coerce to a data.frame):
gar_df <- dbFetch(garibaldi)

```

##### f. `dbplyr` for `dplyr`-like querying!

Let's say we want to do something with a database table, but we really enjoy working with `dplyr` type syntax instead. We'll use the `dbplyr` package to help us out. 

From https://dbplyr.tidyverse.org/: "Note that you don’t actually need to load dbplyr with library(dbplyr); dplyr automatically loads it for you when it sees you working with a database. Database connections are coordinated by the DBI package. Learn more at http://dbi.r-dbi.org/"

```{r}

# Convert to a SQL table:
fish_tbl <- tbl(lter_db, "fish_data")

# Do some wrangling in dplyr format (automatically uses dbplyr): 
yearly_fish <- fish_tbl %>% 
  group_by(year) %>% 
  summarize(
    total_fish = sum(total_count)
  )

# See query:
yearly_fish %>% show_query() # This will show us what the SQL commands would look like! So cool! 

# Actually run the query by passing it to collect(): 
yearly_fish %>% collect()

# Then you don't even need to WRITE any SQL code! Thanks dbplyr!

```

##### g. Adding / removing tables in a .sqlite database

What if we want to add a table (e.g. from a data frame or CSV) to an existing database? 

Use `dbWriteTable` to add a new table to an existing database. For example, let's add the `mtcars` dataset to our database, then remove it:  
```{r}
# View(mtcars)

# Write the table to our database:
dbWriteTable(lter_db, "mtcars_df", mtcars)

# Check to see that it's there: 
DBI::dbListTables(lter_db) # Yup, mtcars_df is now listed! 

# And we can remove it just as easily: 
DBI::dbRemoveTable(lter_db, "mtcars_df")

# Now check: 
DBI::dbListTables(lter_db) # Back to the 3 we started with! 

```

So now you've learned some ways to explore, query and update a database using:

- `DBI` functions to connect & explore
- `RSQLite` as your SQLite engine in R
- SQL (structured query language) to query a database
- `dbplyr` to query a database using `dplyr` syntax
- ...all in R Markdown! 



## End SQL in R introduction
